---
title: "GO analysis"
author: "Ming Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
    keep_md: true
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

## 1. summary

See GO enrichment analysis in Section `3.1` and `3.2`.


## 2. Materials and Methods

**Genes**: here we use `geneList` dataset from [DOSE](https://www.bioconductor.org/packages/DOSE) package.

**Annotation**: Using a Bioconductor package, `org.Hs.eg.db` (version: `3.8.2`)

**Tools**: [clusterProfiler](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3339379/) (R package, `v3.12.0`)


> **GO analysis**
> 
> GO enrichment analysis was performed with the R package clusterProfiler, with a Bonferroni correction and an adjusted p value of 0.01 and qvalue of 0.05 (Yu et al., 2012).
> 
> **Reference**
> 
> 1. Yu G, Wang L-G, Han Y, He Q-Y. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS. 2012;16:284–7. https://doi.org/10.1089/omi.2011.0118.


## 3. Results


> The geneList contains three features:
> 
> numeric vector: fold change or other type of numerical variable
named vector: every number was named by the corresponding gene ID
sorted vector: number should be sorted in decreasing order
> 
> The dataset `geneList` contains three features: 
> 
> 1. numeric vector, foldchange or other type of numerical >variable   
> 2. named vector, gene ID    
> 3. sorted vector, sorted by decreasing order  

```{r}
suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(org.Hs.eg.db))
data(geneList, package = "DOSE")
length(geneList)
head(geneList)
```

Convert the `ENTREZID` to `SYMBOL`

genes with **foldchange > 2** were selected for further analysis (n=239).

```{r}
gene <- names(geneList)[abs(geneList) > 2]
gene.df <- bitr(gene, fromType = "ENTREZID",
        toType = c("ENSEMBL", "SYMBOL"),
        OrgDb = org.Hs.eg.db)

dim(gene.df)
head(gene.df)
```

### 3.1 GO analysis

#### 3.1.1 GO classification 

```{r}
ggo <- groupGO(gene     = gene,
               OrgDb    = org.Hs.eg.db,
               ont      = "CC",
               level    = 3,
               readable = TRUE)

head(ggo)
```


#### 3.1.2 GO over-representation test

Over-representation test(Boyle et al. 2004) were implemented in clusterProfiler. For calculation details and explanation of paramters, please refer to the vignette of DOSE.

```{r}
ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

head(ego)
```

#### 3.1.3 drop specific GO terms or level

Use the function `dropGO()` function to remove specific GO terms or GO level from results of `enrichGO` and `compareCluster`.

```{r, eval = FALSE}
dropGO(ego, level = 4)
dropGO(ego, term = "GO:0140014") # drop the first GO term.
```

#### 3.1.4 test GO at specific level

Use the function `gofilter` to get results at specific GO level from results of `enrichGO` and `compareCluster`.

example: 

```{r, eval = FALSE}
gofilter(ego, level = 4)

```


#### 3.1.5 reduce redundancy of enriched GO terms

Use the function `simplify()` method to calculate semantic similarity among GO terms and remove highly similar terms. Set the cutooff of similarity `cutoff=0.7` and `by=`, `fun=` to decide which column and method to choose between similar temrs.

example: 

Choose the term with minimum `p.adjust` value within the similar GO terms, with cutoff 0.7.

```{r, eval = FALSE}
ego3 <- simplify(ego, cutoff = 0.7, by = "p.adjust", select_fun = min)
```



### 3.2 Make plots


#### 3.2.1 Barplot and dotplot

We could use `barplot()` and `dotplot()` to visualize the results of GO analysis.    

```{r, fig.width=8, fig.width=8}
barplot(ego, showCategory = 15, order = TRUE)
dotplot(ego, showCategory = 15)
```


#### 3.2.2 Gene-Concept Network

Show the linkages of genes and biological concepts as a network.


```{r}
p1 <- cnetplot(ego, foldChange = geneList)
p2 <- cnetplot(ego, foldChange = geneList, categorySize = "pvalue")
p3 <- cnetplot(ego, foldChange = geneList, circular = TRUE, colorEdge = TRUE)

cowplot::plot_grid(p1, p2, p3, nrow = 1, labels = "AUTO", rel_widths = c(.8, .8, 1.2))

```


Or, we can choose custom the labels using `node_label=` parameter.

```{r}
p1 <- cnetplot(ego, node_label = "category")
p2 <- cnetplot(ego, node_label = "gene")
p3 <- cnetplot(ego, node_label = "all")
p4 <- cnetplot(ego, node_label = "none")
cowplot::plot_grid(p1, p2, p3, p4, ncol = 2, labels = "AUTO")
```

#### 3.2.3 Enrichment map

`emapplot()` function: Enrichment map organizes enriched terms into a network with edges connecting overlapping gene sets. In this way, mutually overlapping gene sets are tend to cluster together, making it easy to identify functional module. Support the results from `enrichGO()` and `compareCluster()`.


```{r}
p1 <- emapplot(edo)
p2 <- emapplot(edo, pie_scale=1.5)
p3 <- emapplot(edo, layout="kk")
p4 <- emapplot(edo, pie_scale=1.5, layout="kk") 
cowplot::plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4])
```



```{r}
data(gcSample)
xx <- compareCluster(gcSample, fun="enrichKEGG",
                     organism="hsa", pvalueCutoff=0.05)
p1 <- emapplot(xx)
p2 <- emapplot(xx,legend_n=2) 
p3 <- emapplot(xx,pie="count")
p4 <- emapplot(xx,pie="count", pie_scale=1.5, layout="kk")
cowplot::plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4])
```


## 4. Code

Here are the code for regular GO analysis.

```{r, eval = FALSE}

suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(org.Ce.eg.db))

##

## combine three groups
group_go <- function(gene, orgdb, keytype = "ENTREZID", top = 10) {
  # 
  # gene = g_down
  # orgdb = org.Ce.eg.db
  # keytype = "ENTREZID" 
  # group = "BP"

  # BP
  group <- "BP"
  ggo1 <- groupGO(gene     = gene,
                  OrgDb    = orgdb,
                  keyType  = keytype,
                  ont      = group,  
                  level    = 3,
                  readable = TRUE)
  
  # convert to data.frame
  df1 <- data.frame(ggo1) %>%
    tidyr::unite("name", c("Description", "ID"), sep = "-", remove = FALSE) %>%
    dplyr::arrange(desc(Count)) %>%
    dplyr::mutate(group = group,
                  des = as.character(Description)) %>%
    dplyr::select(name, Count, group, des) %>%
    head(top)
  
  # CC
  group <- "CC"
  ggo2 <- groupGO(gene     = gene,
                  OrgDb    = orgdb,
                  keyType  = keytype,
                  ont      = group,  
                  level    = 3,
                  readable = TRUE)
  
  # convert to data.frame
  df2 <- data.frame(ggo2) %>%
    tidyr::unite("name", c("Description", "ID"), sep = "-", remove = FALSE) %>%
    dplyr::arrange(desc(Count)) %>%
    dplyr::mutate(group = group,
                  des = as.character(Description)) %>%
    dplyr::select(name, Count, group, des) %>%
    head(top)
  
  # MF
  group <- "MF"
  ggo3 <- groupGO(gene     = gene,
                  OrgDb    = orgdb,
                  keyType  = keytype,
                  ont      = group,  
                  level    = 3,
                  readable = TRUE)
  
  # convert to data.frame
  df3 <- data.frame(ggo2) %>%
    tidyr::unite("name", c("Description", "ID"), sep = "-", remove = FALSE) %>%
    dplyr::arrange(desc(Count)) %>%
    dplyr::mutate(group = group,
                  des = as.character(Description)) %>%
    dplyr::select(name, Count, group, des) %>%
    head(top)
  
  ## combine 3 plots/data
  df <- bind_rows(df1, df2, df3) %>%
    tibble::rowid_to_column("x")
  
  # plot
  ggplotColours <- function(n = 6, h = c(0, 360) + 15){
    if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
    hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
  }
  
  colors <- ggplotColours(n = 3)
  colors_axis <- rep(colors, each = top)
  
  p <- ggplot(df, aes(x, Count, fill = group)) +
    geom_bar(stat = "identity") + 
    ggtitle("GO classification") +
    xlab(NULL) + ylab("Number of genes") +
    scale_x_continuous(breaks = df$x, labels = df$des) + 
    scale_fill_manual(values = colors)+
    theme_bw() +
    theme(
      plot.margin = margin(1, 1, 1, 3, "cm"),
      axis.text.x = element_text(angle = 60, hjust = 1, colour = colors_axis),
      legend.position = "top"
    )
  
  return(list(p, df))
}


enrich_go <- function(gene, orgdb, 
                      keytyp = "ENTREZID", 
                      top = 10, 
                      method = "BH", 
                      semsim = TRUE) {
  ## BP
  group <- "BP"
  ego1 <- enrichGO(gene          = gene,
                   OrgDb         = orgdb,
                   ont           = group,
                   pAdjustMethod = method, 
                   pvalueCutoff  = 0.01,
                   qvalueCutoff  = 0.05,
                   readable      = TRUE)
  
  if(isTRUE(semsim)) {
    ego1 <- clusterProfiler::simplify(ego1, cutoff = 0.7, by = "p.adjust", select_fun = min)
  }
  
  df1 <- data.frame(ego1, stringsAsFactors = FALSE) %>%
    tidyr::unite("name", c("Description", "ID"), sep = "-", remove = FALSE) %>%
    dplyr::mutate(des = Description,
                  group = group,
                  plog10 = -log10(p.adjust),
                  qlog10 = -log10(qvalue)) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::select(des, name, Count, plog10, qlog10, group) %>%
    head(top)
    
  ## CC
  group <- "CC"
  ego2 <- enrichGO(gene          = gene,
                   OrgDb         = orgdb,
                   ont           = group,
                   pAdjustMethod = method, 
                   pvalueCutoff  = 0.01,
                   qvalueCutoff  = 0.05,
                   readable      = TRUE)
  
  if(isTRUE(semsim)) {
    ego2 <- clusterProfiler::simplify(ego2, cutoff = 0.7, by = "p.adjust", select_fun = min)
  }
  
  df2 <- data.frame(ego2, stringsAsFactors = FALSE) %>%
    tidyr::unite("name", c("Description", "ID"), sep = "-", remove = FALSE) %>%
    dplyr::mutate(des = Description,
                  group = group,
                  group = group,
                  plog10 = -log10(p.adjust),
                  qlog10 = -log10(qvalue)) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::select(des, name, Count, plog10, qlog10, group) %>%
    head(top)  

  ## MF
  group <- "MF"
  ego3 <- enrichGO(gene          = gene,
                   OrgDb         = orgdb,
                   ont           = group,
                   pAdjustMethod = method, # bonferroni
                   pvalueCutoff  = 0.01,
                   qvalueCutoff  = 0.05,
                   readable      = TRUE)
  
  if(isTRUE(semsim)) {
    ego3 <- clusterProfiler::simplify(ego3, cutoff = 0.7, by = "p.adjust", select_fun = min)
  }
  
  df3 <- data.frame(ego3, stringsAsFactors = FALSE) %>%
    tidyr::unite("name", c("Description", "ID"), sep = "-", remove = FALSE) %>%
    dplyr::mutate(des = Description,
                  group = group,
                  plog10 = -log10(p.adjust),
                  qlog10 = -log10(qvalue)) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::select(des, name, Count, plog10, qlog10, group) %>%
    head(top)
  
  ## combine 3 plots/data
  df <- bind_rows(df1, df2, df3) %>%
    tibble::rowid_to_column("x")
  
  # plot
  ggplotColours <- function(n = 6, h = c(0, 360) + 15){
    if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
    hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
  }
  
  colors <- ggplotColours(n = 3)
  colors_axis <- rep(colors, each = top)
  
  p1 <- ggplot(df, aes(x, plog10, fill = group)) +
    geom_bar(stat = "identity") + 
    ggtitle("GO Enrichment") +
    xlab(NULL) + ylab("-log10(P-value)") +
    scale_x_continuous(breaks = df$x, labels = df$des) + 
    scale_fill_manual(values = colors)+
    theme_bw() +
    theme(
      plot.margin = margin(1, 1, 1, 3, "cm"),
      axis.text.x = element_text(angle = 60, hjust = 1, colour = colors_axis),
      legend.position = "top"
    )
  
  
  labels=function(x) str_wrap(x, width=text_width)
  
  ##
  p2 <- ggplot(df, aes(x, plog10, fill = group)) +
    geom_bar(stat = "identity") + 
    coord_flip() + 
    ggtitle("GO Enrichment") +
    xlab(NULL) + ylab("-log10(P-value)") +
    scale_x_reverse(breaks = df$x, 
                    labels = function(x) stringr::str_wrap(df$des, 40)) + 
    # scale_x_continuous(breaks = rev(df$x), labels = rev(df$des)) + 
    scale_fill_manual(values = colors) +
    theme_bw() +
    theme(
      plot.margin = margin(1, 1, 1, 3, "cm"),
      axis.text.y = element_text(angle = 0, hjust = 1, colour = colors_axis),
      legend.position = "right"
    )
  
  ## output
  pdata <- list(p1, p2, df)
  ego  <- list(ego1, ego2, ego3)
  
  return(list(pdata, ego))
}

```







