---
title: "GO analysis"
author: "Ming Wang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
    keep_md: true
---

## 1. summary

See GO enrichment analysis in Section `3.1` and `3.2`.


## 2. Materials and Methods

**Genes**: Two lists of genes, up-regulated (n=345), down-regulated (n=898).

**Organism**: C. elegans 

**Annotation**: Using a Bioconductor package, `org.Ce.eg.db` (version: `3.8.2`)

**Tools**: [clusterProfiler](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3339379/) (R package, `v3.12.0`)


### GO analysis


GO enrichment analysis was performed with the R package clusterProfiler, with a Bonferroni correction and an adjusted p value of 0.01 and qvalue of 0.05 (Yu et al., 2012).

**Reference**

1. Yu G, Wang L-G, Han Y, He Q-Y. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS. 2012;16:284–7. https://doi.org/10.1089/omi.2011.0118.


## 3. Results

```{r global_options, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 12, 
  fig.height = 8, 
  fig.path = 'Figures/',
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE)
```


```{r, echo = FALSE}
suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(org.Ce.eg.db))
source("go_functions.R")

##
f <- "../../data/Gene list for GO.xlsx"

## up
df_up <- readxl::read_xlsx(f, "UP")
gene1 <- df_up$`Wormbase Identifier`

## down
df_down <- readxl::read_xlsx(f, "DOWN")
gene2   <- df_down$`Wormbase Identifier`

## convert genes to ENTREZID
g1 <- bitr(df_up$`Wormbase Identifier`, "WORMBASE", "ENTREZID", org.Ce.eg.db, drop = TRUE)
g2 <- bitr(df_down$`Wormbase Identifier`, "WORMBASE", "ENTREZID", org.Ce.eg.db, drop = TRUE)

g_up   <- g1$ENTREZID
g_down <- g2$ENTREZID
```


### 3.1 Down-regulated genes (n=898) 

+ GO classification

Gene classification based on GO distribution at `level=3`. 

```{r, echo = FALSE, fig.width=8, fig.height=6}
tmp1 <- group_go(g_down, org.Ce.eg.db)
print(tmp1[[1]]) # plot
```

+ GO enrichment analysis

```{r, echo = FALSE, fig.width=8, fig.height=8}
tmp2 <- enrich_go(g_down, org.Ce.eg.db, method = "bonferroni")
plist <- tmp2[[1]]
dlist <- tmp2[[2]]

# save data
print(plist[[1]])
```

Create the barplot like Figure 5D of Merkwirth et al., 2016 Cell.

```{r, echo = FALSE, fig.width=8, fig.height=8}
print(plist[[2]])
```

Saving above plots to to file: `results/go_enrich.down.pdf`

```{r, echo = FALSE}
pdfFile1 <- "results/go_enrich.down.pdf"

pdf(pdfFile1, width = 8, height = 8)
print(plist[[1]])
print(plist[[2]])
dev.off()

```

Save the original enrichment results to file: `results/go_enrich.down.xlsx`

```{r, echo = FALSE}
if(! dir.exists("results")) dir.create("results")
xlsxFile <- "results/go_enrich.down.xlsx"
xlsx::write.xlsx(dlist[[1]], xlsxFile, "BP")
xlsx::write.xlsx(dlist[[2]], xlsxFile, "CC", append = TRUE)
xlsx::write.xlsx(dlist[[3]], xlsxFile, "MF", append= TRUE)
```

### 3.2 Up-regulated genes (n=345) 

+ GO classification

```{r, echo = FALSE, fig.width=8, fig.height=6}
tmp3 <- group_go(g_up, org.Ce.eg.db)
print(tmp3[[1]]) # plot
```

+ GO enrichment analysis

```{r, echo = FALSE, fig.width=8, fig.height=8}
tmp4 <- enrich_go(g_down, org.Ce.eg.db, method = "bonferroni")
plist <- tmp4[[1]]
dlist <- tmp4[[2]]

# save data
print(plist[[1]])
```

Create the barplot like Figure 5D of Merkwirth et al., 2016 Cell.

```{r, echo = FALSE, fig.width=8, fig.height=8}
print(plist[[2]])
```

Saving above plots to to file: `results/go_enrich.up.pdf`

```{r, echo = FALSE}
pdfFile1 <- "results/go_enrich.up.pdf"

pdf(pdfFile1, width = 8, height = 8)
print(plist[[1]])
print(plist[[2]])
dev.off()

```



Save the original enrichment results to file: `results/go_enrich.up.xlsx`

```{r, echo = FALSE}
if(! dir.exists("results")) dir.create("results")
xlsxFile <- "results/go_enrich.up.xlsx"
xlsx::write.xlsx(dlist[[1]], xlsxFile, "BP")
xlsx::write.xlsx(dlist[[2]], xlsxFile, "CC", append = TRUE)
xlsx::write.xlsx(dlist[[3]], xlsxFile, "MF", append= TRUE)
```


